flowCore: A Bioconductor software package for high throughput flow cytometry data analysis

### 2008-11-21 paper/bmc_article.tex
flowCore to BMC bioinfo
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35431 bc3139a8-67e5-0310-9ffc-ced21a209358

\documentclass[12pt]{article}
\usepackage{setspace}
\usepackage{url}
\usepackage[authoryear,round]{natbib}
\usepackage{graphicx} 
\newcommand{\Rpackage}[1]{{\textsf{#1}}}
\newcommand{\Rclass}[1]{{\textit{#1}}}

\title{flowCore: A Bioconductor software package for high throughput
  flow cytometry data analysis \footnote{This work was supported by
    NIH grant EB005034 and by the Michael Smith Foundation for Health
    Research. RRB is an ISAC Scholar.}}

\author{Florian Hahne (\textit{Fred Hutchinson Cancer Research
    Center})\\
  Nolwenn LeMeur* (\textit{Fred Hutchinson Cancer Research Center})\\
  Ryan R. Brinkman (\textit{BC Cancer Research Centre})\\
  Byron Ellis (\textit{AdBrite, Inc})\\
  Perry Haaland (\textit{BD Biosciences})\\
  Deepayan Sarkar (\textit{Fred Hutchinson Cancer Research
    Center})\\
  Josef Spidlen (\textit{BC Cancer Research Centre})\\
  Errol Strain (\textit{BD Biosciences})\\
  Robert Gentleman (\textit{Fred Hutchinson Cancer Research
    Center})\\
}

\doublespacing
\begin{document}
\begin{singlespace}
\maketitle

Corresponding author:\\
\textit{
  Florian Hahne, PhD\\
  Computational Biology Program\\
  Division of Public Health Sciences\\
  Fred Hutchinson Cancer Research Center\\
  1100 Fairview Ave. N, M2-B876\\
  PO Box 19024\\
  Seattle, Washington 98109-1024\\
  phone: 206-667-3148\\
  fax: 206-667-1319\\
  fhahne@fhcrc.org
}

\end{singlespace}

\section*{Abstract}
\subsubsection*{Background}
Recent advances in automation technologies have enabled the use of
flow-cytometry high content screening (FH-HCS). However, data
management and data analysis methods have not advanced sufficiently
far from the inital small-scale studies to support the large complex
data sets generated in FH-HCS applications.
\subsubsection*{Methods}
In this manuscript, we describe a set of flexible
computational tools for FC-HCS data analysis written 
in R and freely available from the Bioconductor Project.

\subsubsection*{Results}
The \Rpackage{flowCore} package provides the flow cytometry community
with a shared research platform that enables bioinformaticians,
computer scientists, and statisticians to work collaboratively with
biologists and clinicians to develop novel methods for FCM data
analysis.
\subsubsection*{Conclusions}
We hope that our framework will be the foundation for fruitful 
collaborative research involving multiple scientific fields and
that it will help facilitate the
development and deployment of FC-HCS methods to complex 
scientific and clinical applications.

\section*{Introduction}
\subsection*{Computational tools for the analysis of FC-HCS data}
Automation technologies developed during the last several years have
enabled the use of FCM high content screening (FC-HCS) to generate
large, complex datasets in both basic and clinical research
applications \citep{brinkman2007hcf}. A serious bottleneck in the
interpretation of existing studies and the application of FC-HCS to
even larger, more complex problems is that data management and data
analysis methods have not advanced sufficiently far from the methods
developed for applications of flow cytometry (FCM) to small-scale,
tube-based studies.

@@

\section*{Materials and Methods}
\subsection*{Basic data structures}
\subsubsection*{flowFrame}

@@

%%Figure1 Diagram of the data structures and the basic operations
\begin{singlespace}
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{For each experiment, the content of
    the FCS files, phenotypic and metadata are stored in a
    \Rclass{flowSet}. Each \Rclass{flowFrame} in a \Rclass{flowSet}
    corresponds to one FCS file. All basic operations (e.g.,
    compensation, transformation, gating) can be applied to either
    single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.}}
\end{figure}
\end{singlespace}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Data Transformations} \\
\hline
linear & $ax + b$ \\
quadratic & $ax^2 + bx + c$ \\
natural logarithm & $ln(x)(r/d)$ \\
logarithm & $log(x)(r/d)$ \\
biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
truncate & $x_{x<=a} = a$ \\
scale & $(x-a)/(b-a)$ \\
split-scale & combination of linear and logarithm \\
arcsinh & $asinh(a + bx)+c$ \\
\hline
\end{tabular}
\caption{\label{table1}Data transformations implemented in
  \Rpackage{flowCore}. Within these formulas, $x$ is the variable
  corresponding to value being transformed, $y$ is the result of the
  transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
  constants affecting the transformation function, $e$ is the base of
  the natural logarithm, and $root$ is a root-finding algorithm.  }
\end{center}
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Gates} \\
\hline
rectangleGate & n-dimensional rectangular regions \\
quadGate & quadrant regions in two dimensions \\
polygonGate & polygonal regions in two dimensions \\
polytopeGate & generalization of polygon in n dimensions \\
ellipsoidGate & n-dimensional ellipsoid region \\
\hline
\multicolumn{2}{|c|}{Filters} \\
\hline
sampleFilter & random sub-sampling\\
expressionFilter & results of a boolean expression \\
kmeansFilter & K-means clustering \\
norm2Filter & bivariate normal distribution \\
curv1Filter & local density regions in 1D \\
curv2Filter & density regions in 2D \\
timeFilter & abnormal data acquisition over time \\
\hline
filterSet & gating strategies \\
\hline
\end{tabular}
\caption{\label{table2}Filter and gate classes implemented in
  \Rpackage{flowCore}.}
\end{center}
\end{table}

\section*{Results}
The \Rpackage{flowCore} package has been successfully applied in the
analysis of several datasets
\citep{gasparetto2004ice,brinkman2007hcf}. A complete documentation of
the \Rpackage{flowCore} software is beyond the scope of this
publication.  Much more comprehensive documentation and users guide
information with programmatic examples is available online
(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
and as part of the package distribution. Here, we want to briefly
exemplify some of the software's key features, that is, the coherent
treatment of all samples in a potentially large experiment, the
concept of data-driven automated gating, the integration of existing
software into the framework, and the generation of publication-quality
graphics for data visualization.

@@

\begin{singlespace}
\begin{figure}[htbp]
\centering
\includegraphics[width=0.98\textwidth]{flowQ.jpg}
\caption{\label{flowQ}%
  HTML quality assessment report generated by the flowQ package for a
  subset of data from an experiment focussing on Graft-Versus-Host
  Disease \citep{brinkman2007hcf}. Rows correspond to the samples in
  the set, columns to different quality checks.}
\end{figure}
\end{singlespace}

@@

\begin{singlespace}
\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
  Scatterplot matrix of a single \Rclass{flowSet} from an experiement
  focussing on immune tolerance following kidney
  transplantation. Outlines of the gating regions identified by an
  automated gating operation are added on top of the density
  representation of the data.}
\end{figure}
\end{singlespace}

\section*{Discussion}

Through \Rpackage{flowCore}, we have provided the FCM community with
an open source, freely available, highly functional, standards
compliant development and analysis platform for high throughput data
analysis.  Our intention is to provide a platform for the
collaborative development of new analysis methods that will facilitate
the transition of these new methods to the larger flow community.  Our
experience has been that the collaborative effort we enable for
devising new methodology has been proven beneficial for a number of
different biological and computational biology challenges.  We hope
that our framework will be the foundation for fruitful shared research
by many collaborators from multiple scientific fields and will help
resolve bottlenecks that currently prevent further development and
deployment of FC-HCS to increasingly complex and important scientific
and clinical applications.

\section*{Acknowledgements}
This work was supported by NIH grant EB005034 and by the Michael Smith
Foundation for Health Research. RRB is an ISAC Scholar.

\bibliographystyle{plainnat}  
\bibliography{flowCoreRef} 
\end{document}

>>>>

%% BioMed_Central_Tex_Template_v1.05
%%                                      %
%  bmc_article.tex            ver: 1.05 %
%                                       %

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%         <27 January 2006>           %%
%%                                     %%
%%                                     %%
%% Uses:                               %%
%% cite.sty, url.sty, bmc_article.cls  %%
%% ifthen.sty. multicol.sty		       %%
%%									   %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%	
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.pdf and the instructions for    %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\NeedsTeXFormat{LaTeX2e}[1995/12/01]
\documentclass[10pt]{bmc_article}

% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%   
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %% 
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %% 
%%  submitted article.                         %% 
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}

\newboolean{publ}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                              %%
%% You may change the following style settings  %%
%% Should you wish to format your article       %%
%% in a publication style for printing out and  %%
%% sharing with colleagues, but ensure that     %%
%% before submitting to BMC that the style is   %%
%% returned to the Review style setting.        %%
%%                                              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Review style settings
\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

% Begin ...
\begin{document}
\begin{bmcformat}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{flowCore: A Bioconductor software package for high throughput
  flow cytometry data analysis}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Ensure \and is entered between all but   %%
%% the last two authors. This will be       %%
%% replaced by a comma in the final article %%
%%                                          %%
%% Ensure there are no trailing spaces at   %% 
%% the ends of the lines                    %%     	
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Florian Hahne\correspondingauthor$^{1}$%
       \email{Florian Hahne\correspondingauthor - fhahne@fhcrc.org}%
      \and
         Nolwenn LeMeur\correspondingauthor$^{1,2}$%
         \email{Nolwenn LeMeur\correspondingauthor - nlemeur@irisa.fr}
       and 
         Ryan R. Brinkman$^3$%
         \email{Ryan R. Brinkman - rbrinkman@bccrc.ca}%
       and
          Byron Ellis$^4$%
          \email{Byron Ellis - byron.ellis@gmail.com}
       and
          Perry Haaland$^5$%
          \email{Perry Haaland - Perry_D_Haaland@bd.com}
       and
          Deepayan Sarkar$^{1}$%
          \email{Deepayan Sarkar - dsarkar@fhcrc.org}
       and
         Josef Spidlen$^3$%
         \email{Josef Spidlen - jspidlen@bccrc.ca}
       and
         Errol Strain$^5$%
         \email{Errol Strain - Errol_Strain@bd.com  }  
       and
          Robert Gentleman$^{1}$%   
          \email{Robert Gentleman - rgentlem@fhcrc.org}
      }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address{%
    \iid(1)Life Sciences Department, Computational Biology Program,%
    Division of Public Health Sciences, Fred Hutchinson Cancer Research Center,% 
    1100 Fairview Ave. N, M2-B876, PO Box 19024,  Seattle, Washington 98109-1024\\
    \iid(2)INSERM, IRISA - Symbiose, Campus Beaulieu, Université de Rennes I, %
    35042 Rennes Cedex,France \\
    \iid(3) Terry Fox Laboratory, British Columbia Cancer Agency Research Centre, %
    675 West 10th Avenue, Vancouver, BC V5Z 1L3  Canada, Canada\\
    \iid(4) AdBrite Inc\\
    \iid(5) BD Biosciences\\

}%

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the Instructions for       %%
%% authors on http://www.biomedcentral.com  %%
%% and change the section headings          %%
%% accordingly.                             %%   
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
  \paragraph*{Background:} Recent advances in automation technologies
  have enabled the use of flow-cytometry high content screening
  (FH-HCS). However, data management and data analysis methods have
  not advanced sufficiently far from the inital small-scale studies to
  support the large complex data sets generated in FH-HCS
  applications.

  \paragraph*{Results:} The \Rpackage{flowCore} package provides the
  flow cytometry community with a shared research platform that
  enables bioinformaticians, computer scientists, and statisticians to
  work collaboratively with biologists and clinicians to develop novel
  methods for FCM data analysis.

  \paragraph*{Conclusions:} We hope that our framework will be the
  foundation for fruitful collaborative research involving multiple
  scientific fields and that it will help facilitate the development
  and deployment of FC-HCS methods to complex scientific and clinical
  applications.
\end{abstract}

\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the instructions for       %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and change the section headings          %%
%% accordingly.                             %% 
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
 \subsection*{Computational tools for the analysis of FC-HCS data}
Automation technologies developed during the last several years have
enabled the use of FCM high content screening (FC-HCS) to generate
large, complex datasets in both basic and clinical research
applications \citep{brinkman2007hcf}. A serious bottleneck in the
interpretation of existing studies and the application of FC-HCS to
even larger, more complex problems is that data management and data
analysis methods have not advanced sufficiently far from the methods
developed for applications of flow cytometry (FCM) to small-scale,
tube-based studies.

@@

%%%%%%%%%%%%%%%%%%
\section*{Implementation}
 \subsection*{Basic data structures}
\subsubsection*{flowFrame}

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
  \subsection*{From quality assessment to batch gating}
  The \Rpackage{flowCore} package has been successfully applied in the
  analysis of several datasets
  \citep{gasparetto2004ice,brinkman2007hcf}. A complete documentation
  of the \Rpackage{flowCore} software is beyond the scope of this
  publication.  Much more comprehensive documentation and users guide
  information with programmatic examples is available online
  (\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
  and as part of the package distribution. Here, we want to briefly
  exemplify some of the software's key features, that is, the coherent
  treatment of all samples in a potentially large experiment, the
  concept of data-driven automated gating, the integration of existing
  software into the framework, and the generation of
  publication-quality graphics for data visualization.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
  Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Authors contributions}
    Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  This work was supported by NIH grant EB005034 and by the Michael Smith
Foundation for Health Research. RRB is an ISAC Scholar.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%              
%%  Bmc_article.bst  will be used to                       %%
%%  create a .BBL file for submission, which includes      %%
%%  XML structured for BMC.                                %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %% 
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %% 
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{flowCoreRef} }     % Bibliography file (usually '*.bib' )

%%%%%%%%%%%

\ifthenelse{\boolean{publ}}{\end{multicols}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \subsection*{Figure 1 - Sample figure title}
      A short description of the figure content
      should go here.

  \subsection*{Figure 2 - Sample figure title}
      Figure legend text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Sample table title}
    Here is an example of a \emph{small} table in \LaTeX\ using  
    \verb|\tabular{...}|. This is where the description of the table 
    should go. \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        A1 & B2  & C3 \\ \hline
        A2 & ... & .. \\ \hline
        A3 & ..  & .  \\ \hline
      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{bmcformat}
\end{document}
